DCMIP2016, Part 1: Models and Equation Sets

### 2016-12-26 DCMIP2016-Part1-Models.tex
- Added DYNAMICO text - More paper cleanup

@@

The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.  The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.  Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.

\begin{figure}
\begin{center}
\includegraphics[width=2in]{CubedSphereMesh.pdf}
\end{center}
\caption{An example cubed-sphere grid.} \label{fig:CubedSphereMesh}
\end{figure}

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  An example of such a grid is depicted in Figure \ref{fig:ICONMesh}.  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

\begin{figure}
\begin{center}
\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf}
\end{center}
\caption{An icosahedral (triangular) grid with additional refinement over Europe (shown in red).} \label{fig:ICONMesh}
\end{figure}

\subsection{Icosahedral (hexagonal) grid}

\begin{figure}
\begin{center}
\includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
\end{center}
\caption{An icosahedral (hexagonal) grid.} \label{fig:ICONIcosahedlalHexGrid}
\end{figure}

@@

Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.  For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.  However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.  This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.  Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e. for ocean basins or ice sheets) \citep{ringler2008multiresolution}.

\begin{figure}
\begin{center}
\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
\end{center}
\caption{A constrained centroidal Voronoi tessellation with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
\end{figure}

@@

Due to its association with hydrostatic models, it is common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

@@

Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
\begin{align} \label{eq:PrognosticWEquationK2}
\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}_h) \cdot \vb{k}
\end{align}  and
\begin{align} \label{eq:PrognosticUhEquationK2}
\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.

@@

The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
\begin{align} \label{eq:VerticalVelocityOrthogonalS}
\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k}.
\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
\end{align}  Note that the last term of this expression can then be expanded using
\begin{align}
(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.

@@

In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.  This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.  The horizontal covariance operator is defined by
\begin{align}
[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
\end{align} Hence the covariant horizontal velocity is given by
\begin{align}
\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
\end{align}  For a time-dependent $s$ coordinate,
\begin{align}
\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
\end{align} and so, using (\ref{eq:VerticalVelocityOrthogonalS}) and (\ref{eq:PrognosticUhEquationK2S}) the evolution equation for the covariant horizontal velocity reads
\begin{align}
\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
%=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
%& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
%& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [(2 \vg{\Omega} + \vg{\zeta}) \times \vb{u}]_s \nonumber \\
& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) + (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right]. \label{eq:CovariantVelocityUpdateS}
\end{align}  Again, observe that this equation simplifies significantly if $(\partial z / \partial t)_s = w$.

@@

%Then using (\ref{eq:MaterialDerivativeS}) gives
%\begin{align}
%\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{v}_h) 
%\end{align}

@@

@@

\subsubsection{ICON}

@@

{\color{red}In this section provide a short description (approximately 0.5 pages) of the dynamical core, focusing on unique features or design specifications.  Do not include information on the physical parameterizations used by the modeling system.  Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}

@@

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by Arakawa and Konor (2009), which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Time-differencing is based on the third-order Adams-Bashforth  scheme. Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.  
\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}

FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

@@

{\color{red}[HALL]}

@@

{\color{red}[DUBOS]}

>>>>

Model grids are a primary concern for dynamical cores, since they determine how degrees of freedom are distributed across the globe.  In order to meet demands for high computational efficiency and equal partitioning of computation across large parallel systems, modern dynamical cores have been exploring a number of options for model grids.  This section describes many of the model grids that have been employed globally.

@@

The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.  The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.  Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities.  This grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (left).

\begin{figure}
\begin{center}
\includegraphics[width=2in,trim={0cm 4.7cm 0cm 4cm},clip]{CubedSphereMesh.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_nested_schematic.pdf} \quad \includegraphics[width=2.2in, trim={4cm 8.2cm 4cm 8.2cm},clip]{ICON_grid_dual_R2B3_noWorldMap.pdf}
\end{center}
\caption{(Left) A cubed-sphere grid with 16 elements along each cubed-sphere edge.  (Center) An icosahedral (triangular) grid with additional refinement over Europe, as indicated in red.  (Right) An icosahedral (hexagonal) grid.} \label{fig:CubedSphereGeodesicMesh}
\end{figure}

@@

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  An example of such a grid is depicted in Figure \ref{fig:CubedSphereGeodesicMesh} (center).  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

\subsection{Icosahedral (hexagonal) grid / geodesic grid} \label{sec:IcosahedralHexagonalGrid}

The icosahedral (hexagonal) grid, also commonly referred to as the geodesic grid, is obtained by taking the dual to the icosahedral (triangular grid), namely by replacing grid nodes with spherical polygons.  Their cells are hexagonal, except for twelve pentagonal cells.  Given a icosahedral-triangular mesh, vertices of the corresponding icosahedral-hexagonal mesh are defined as either circumcenters or barycenters of triangles, leading to either a Voronoi mesh, used by DYNAMICO, or a barycentric mesh, used by NICAM.  A Voronoi mesh has the property that triangular edges are perpendicular to edges of hexagons/pentagons, facilitating the formulation of certain finite-difference and finite-volume numerical schemes.  The resulting highly homogeneous and isotropic grid then appears analogous to the grid in Figure \ref{fig:CubedSphereGeodesicMesh} (right).  Unlike the cubed-sphere and icosahedral (triangular) grid, grid cells on this the geodesic grid are guaranteed to be edge-neighbors (cells that share a given edge) if they are also node-neighbors (cells that share a given node).

It is useful to optimize icosahedral-hexagonal grids as well. DYNAMICO applies a number of iterations of Lloyds algorithm \citep{lloyd1982least}, following which one replaces the vertices of the original triangular mesh by the centroid of hexagons/pentagons, then re- generates the icosahedral-hexagonal mesh. This improves the homogeneity of the grid (e.g. ratio of largest cell area to smallest cell area) but several thousand iterations may be required for a significant improvement.

@@

Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.  For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.  However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.  This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.  Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e. for ocean basins or ice sheets) \citep{ringler2008multiresolution}.  CCVT grids are often confused with deformations of the icosahedral (hexagonal) grid described in section \ref{sec:IcosahedralHexagonalGrid}, since both typically contain a large number of hexagonal elements.  However, CCVT grids are fundamentally constructed using a very different technique and will often contain polygonal elements with more than six sides.

\begin{figure}
\begin{center}
\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
\end{center}
\caption{A constrained centroidal Voronoi tessellation grid with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
\end{figure}

@@

It is also common to use the 2D Kinetic energy, $K_2 = \frac{1}{2} (\vb{u}_h \cdot \vb{u}_h)$.  Decomposing the momentum transport term into horizontal and vertical components gives
\begin{align}
\vb{u} \cdot \nabla \vb{u} =& \vb{u}_h \cdot \nabla \vb{u}_h + (\nabla \times \vb{u}_h) \times (w \vb{k}) + (\vb{u} \cdot \nabla w) \vb{k}.
\end{align}  The first term in this expression admits the relationships
\begin{align}
\left[ \vb{u}_h \cdot \nabla \vb{u}_h \right]_z =& \nabla_z K_2 + \zeta_h \vb{k} \times \vb{u}_h, \\
(\vb{u}_h \cdot \nabla \vb{u}_h) \cdot \vb{k} =& - \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) = - K_2 (\nabla \cdot \vb{k}) - \tfrac{1}{2} \left[ \vb{u}_h \cdot (\nabla \times \vb{u}_t)  + (\nabla \times \vb{u}_h) \cdot \vb{u}_t \right]
\end{align} where $\zeta_h = (\nabla \times \vb{u}) \cdot \vb{k} = (\nabla \times \vb{u}_h) \cdot \vb{k}$ is the relative vorticity scalar and $\vb{u}_t = \vb{k} \times \vb{u}_h$.  Note that this equation does incorporate metric terms associated with horizontal advection of $\vb{k}$ which must be accounted for.

@@

Thus the vertical velocity equation, obtained by taking (\ref{eq:PrognosticVelocity1}) $\cdot \vb{k}$, leading to
\begin{align} \label{eq:PrognosticWEquationK2}
\pdiff{w}{t} =& \vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k}) - \vb{u} \cdot \nabla w - \frac{1}{\rho} \pdiff{p}{z} - \pdiff{\Phi}{z} - (2 \vg{\Omega} \times \vb{u}) \cdot \vb{k}
\end{align}  and
\begin{align} \label{eq:PrognosticUhEquationK2}
\pdiff{\vb{u}_h}{t} =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - (\nabla \times \vb{u}_h) \times (w \vb{k}) - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z.
\end{align}  Note that under the shallow atmosphere approximation, the metric term $\vb{u}_h \cdot (\vb{u}_h \cdot \nabla \vb{k})$ in (\ref{eq:PrognosticWEquationK2}) is set equal to zero in accordance with \cite{phillips1966equations}.

@@

The prognostic equation for vertical velocity from (\ref{eq:VerticalVelocityOrthogonal}) is then reformulated using (\ref{eq:ChainRule}),
\begin{align} \label{eq:VerticalVelocityOrthogonalS}
\left(\pdiff{w}{t} \right)_s =& \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K + \Phi) \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}_h \right] \cdot \vb{k}.
\end{align}  Similarly, the prognostic equation for horizontal velocity from (\ref{eq:HorizontalVelocityOrthogonal}) is reformulated as
\begin{align} \label{eq:HorizontalVelocityOrthogonalS}
\left( \pdiff{\vb{u}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_z p - \nabla_z (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_z + \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{u}_h).
\end{align}  Note that the last term of this expression can then be expanded using
\begin{align}
(\vb{k} \cdot \nabla \vb{u}_h) =& - \vb{u}_h \cdot \nabla \vb{k} - \vb{k} \times (\nabla \times \vb{u}_h),
\end{align} where under the shallow-atmosphere approximation $(\vb{u}_h \cdot \nabla \vb{k})$ is typically dropped.

@@

In conjunction with (\ref{eq:HorizontalChainRule}), the horizontal momentum equation (either  (\ref{eq:HorizontalVelocityOrthogonal}) or (\ref{eq:PrognosticUhEquationK2})) with an arbitrary vertical coordinate gives rise to a two-term pressure gradient.  This can be avoided by prognosing the covariant components of the velocity in place of the physical velocity components.  The horizontal partial covariance operator is defined by
\begin{align}
[\vb{b}]_s \equiv [\vb{b}]_z + (\nabla_s z) (\vb{k} \cdot \vb{b}).
\end{align} Hence the covariant horizontal velocity is given by
\begin{align}
\vb{v}_h \equiv [\vb{u}]_s = \vb{u}_h + (\nabla_s z) w.
\end{align}  For a time-dependent $s$ coordinate,
\begin{align}
\left[ \pdiff{}{t} (\nabla_s z) \right]_s =& \nabla_s \left( \pdiff{z}{t} \right)_s,
\end{align} and so, using (\ref{eq:VerticalVelocityOrthogonalS}) and (\ref{eq:PrognosticUhEquationK2S}) the evolution equation for the covariant horizontal velocity reads
\begin{align}
\left( \pdiff{\vb{v}_h}{t} \right)_s =& \left( \pdiff{\vb{u}_h}{t} \right)_s + \left( \pdiff{w}{t} \right)_s (\nabla_s z) + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
=& - \frac{1}{\rho} \nabla_z p - \nabla_z (K_2 + \Phi) - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \vb{k} \times (\nabla \times \vb{u}_h)  - \zeta_h \vb{k} \times \vb{u}_h - [2 \vg{\Omega} \times \vb{u}]_z - w (\vb{u}_h \cdot \nabla \vb{k}) \\
& + (\nabla_s z) \left\{ \pdiff{s}{z} \left[ \left( \pdiff{z}{t} \right)_s \pdiff{w}{s} - \frac{1}{\rho} \pdiff{p}{s} - \pdiff{}{s} (K_2 + \Phi) - w \pdiff{w}{s} \right] - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right] \cdot \vb{k} \right\} \\
& + w \nabla_s \left( \pdiff{z}{t} \right)_s \\
=& - \frac{1}{\rho} \nabla_s p - \nabla_s (K_2 + \Phi) + w \nabla_s \left( \pdiff{z}{t} \right)_s - w (\vb{u}_h \cdot \nabla \vb{k}) - [( \vg{\zeta} + 2 \vg{\Omega}) \times \vb{u}]_s \nonumber \\
& \qquad - \left[ \left( \pdiff{z}{t} \right)_s - w \right] \left[ \vb{k} \times (\nabla \times \vb{u}_h) - (\nabla_s z) \left( \pdiff{s}{z} \right) \pdiff{w}{s} \right]. \label{eq:CovariantVelocityUpdateS}
\end{align}  Again, observe that this equation simplifies significantly if $(\partial z / \partial t)_s = w$.

@@

%In the case of an Eulerian formulation with $(\partial z / \partial t)_s = 0$, we can use
%\begin{align}
%0 = \nabla (\vb{k} \cdot \vb{u}_h) = (\vb{k} \cdot \nabla \vb{u}_h) + (\vb{u}_h \cdot \nabla \vb{k}) + \vb{k} \times (\nabla \times \vb{u}_h)
%\end{align} to obtain
%\begin{align}
%\left( \pdiff{\vb{v}_h}{t} \right)_s =& - \frac{1}{\rho} \nabla_s p - \nabla_s (K + \Phi) - \left[ (\vg{\zeta} + 2 \vg{\Omega}) \times \vb{u} \right]_s - w \left( \pdiff{s}{z} \right)^{-1} \nabla_s \left( \pdiff{s}{t} \right) - \left( \pdiff{z}{t} \right)_s (\vb{k} \cdot \nabla \vb{v}_h) 
%\end{align}

@@

\subsubsection{DYNAMICO} \label{sec:DYNAMICOEquations}

DYNAMICO is a shallow-atmosphere fully-compressible model based on a mass-based vertical coordinate and a Hamiltonian formulation (Dubos and Tort, 2014). Prognostic variables are pseudo-density ?s, mass-weighted tracers (potential temperature, water species), geopotential $\Phi$, ``horizontal'' covariant components of momentum and mass-weighted vertical momentum $W = \rho_s g^{-2} d\Phi/dt = \rho_s g^{-1} w$. Prognostic equations are in flux-form for mass, tracers and $W$, in advective form for $\Phi$ and in vector-invariant form for ``horizontal'' momentum.

@@

DYNAMICO uses an additive Runge-Kutta time scheme with two Butcher tableaus, one explicit and one implicit. A Hamiltonian splitting decides which terms of the equations of motion are treated explicitly or implicitly (Dubos and Dubey, in preparation). As a result the implicit terms couple the vertical acceleration due to the pressure gradient and the adiabatic pressure change due to vertical displacements of fluid parcels. The resulting implicit problem reduces to independent, scalar, purely vertical, nonlinear problems which are solved to machine precision in two Newton iterations involving one tridiagonal solve each. The overall time scheme has a HEVI (horizontally explicit, vertically implicit) structure. Currently the second-order 3-stage scheme ARK(2,3,2) is used \citep{giraldo2013implicit}.

@@

This section provide a short description of unique features or design specifications from several modern dynamical cores.

%Do not include information on the physical parameterizations used by the modeling system.  Make reference to the model grid employed from section \ref{sec:ModelGrids}, the specific equation set being discretized by the model in section \ref{sec:EquationSets}, explicit numerical techniques for diffusion and stabilization in section \ref{sec:DiffusionStabilization}, filters and fixers in section \ref{sec:FiltersFixers} and the temporal discretization in section \ref{sec:TemporalDiscretizations}.}

@@

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by Arakawa and Konor (2009), which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Time-differencing is based on the third-order Adams-Bashforth  scheme. Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.  
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry \citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using builing blocks from the litterature \citep{Thuburn2009Numerical,Ringler2010Unified}, and yielding, to a certain extent, an energy-conserving spatial discretization
\item A novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

\subsection{FV Cubed (FV$^3$)}

The Geophysical Fluid Dynamics Laboratory FV$^3$ model uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

@@

@@
